preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-08 Manuscript/manuscript.Rmd
Discussion edits

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on ground truth TAD/loop boundaries obtained from a TAD/loop calling tools and ChIP-seq defined genomic annotations as the predictor space. We benchmarked *preciseTAD* against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from both GM12878 and K562. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and re-sampling techniques. We show that genomic distance between boundary regions and annotations, coupled with random under-sampling help to form the predictor space resulting in optimal model performance. <!--We demonstrate that binding of four transcription factors (SMC3, Rad21, CTCF, Znf143) are sufficient for accurate TAD and loop boundary predictions. The model predicts probability for each base to be a baoundary, and, coupled with spatial clustering (DBSCAN) and scalable partition around medoids clustering (CLARA), allows for the detection of boundaries at the base-level resolution.-->

Our machine learning framework yielded several interesting observations. First, we demonstrated that RF models built using *distance*-type predictors outperformed models built on previously published feature engineering techniques including signal strength, overlap counts and overlap percents [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa; @kai2018predicting]. We further demonstrated that class imbalance hinders boundary prediction, but can be effectively addressed by a simple random under-sampling technique, an aspect of boundary prediction that was unaddressed in previous studies [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa]. We showed that information about only four transcription factors (SMC3, Rad21, CTCF, Znf143) is necessary and sufficient for accurate boundary prediction, outperforming histone modification- and BroadHMM-built models [@Moore:2015aa; @Al-Bkhetan:2018aa]. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@hansen2018recent; @sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Interestingly, the same transcription factors accurately predicted both TAD and loop boundaries, suggesting similarity of the mechanisms of TAD and loop formation. <!--Although our observations contradict a previous study showing histone marks alone were effective in predicting TAD boundaries [@huang2015predicting], our findings are backed by a large body of literature [??? Add REFs]. -->This suggested that the random forest model, when tuned and feature engineered correctly, is highly effective in predicting domain boundaries.

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC). Our aim was to have a balanced metric that was sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values [??? So why didn't you use it?]. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling.

Guided by both low-resolution Hi-C data and high-resolution genome annotation data, *preciseTAD* predicts base-level resolution boundaries, alleviating resolution limitations of Hi-C data. We show that, unlike Arrowhead, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* TAD and loop boundaries were enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143, supporting recent observations that TADs and loops are dynamic structures generated by similar mechanisms [@hansen2018recent]. Likewise *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, a known feature among the 3D architecture of the human genome, further highlighting their biological significance [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa].

*preciseTAD* offers flexibility in controlling both the number of TAD(loop) boundaries as well as the distance between them. The two primary parameters are the probability threshold (*t*) and $\epsilon$ (refered to as eps; parameter of DBSCAN). The combination of these two quantities change the resulting number of predicted boundaries from *preciseTAD*. A lower threshold and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 4 different thresholds ($t=(0.95,0.975,0.99,1.0)$) and 4 different $\epsilon$ values ($\epsilon=(0.2 r, 0.5r, r, 2r)$)), where $r=5000$, the optimal resolution. We found that the normalized overlaps--calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries --between top TFBS sites and flanked preciseTAD boundaries were maximized for combinations of t=1.0 and eps=r (Supplementary Figure 9). The same heuristic ($t=1.0$ and $\epsilon=r$) ensured that boundaries were at least 5kb bases away from each other, while also suitably enriched for architectural proteins (Supplementary Figures 10-11).

>>>>

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on ground truth TAD/loop boundaries obtained from a TAD/loop calling tools and genomic annotations as the predictor space. We benchmarked *preciseTAD* against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We applied *preciseTAD* on cell line-specific boundaries and annotation data from both GM12878 and K562. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and re-sampling techniques. We show that genomic distance between high-resolution boundary regions and transcription factor binding sites, coupled with random under-sampling results in best model performance. We demonstrate that binding of four transcription factors (SMC3, Rad21, CTCF, Znf143) are sufficient for accurate TAD and loop boundary predictions. Applied to each annotated base, the model predicts probability for each base to be a boundary, and, coupled with spatial clustering (DBSCAN) and scalable partition around medoids clustering (CLARA), allows for the detection of boundaries at the base-level resolution. *preciseTAD* is implemented as an R package, and allows for precise prediction of boundaries for any cell type for which the corresponding genomic annotations are available.

Our machine learning framework yielded several interesting observations. First, we demonstrated that RF models built using *distance*-type predictors outperformed models built on previously published feature engineering techniques including signal strength, overlap counts and overlap percents [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa; @kai2018predicting]. We further demonstrated that class imbalance hinders boundary prediction, but can be effectively addressed by a simple random under-sampling technique, an aspect of boundary prediction that was unaddressed in previous studies [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa]. We showed that information about only four transcription factors (SMC3, Rad21, CTCF, Znf143) is necessary and sufficient for accurate boundary prediction, outperforming histone modification- and BroadHMM-built models [@Moore:2015aa; @Al-Bkhetan:2018aa]. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@hansen2018recent; @sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Interestingly, the same transcription factors accurately predicted both TAD and loop boundaries, suggesting similarity of the mechanisms of TAD and loop formation. <!--Although our observations contradict a previous study showing histone marks alone were effective in predicting TAD boundaries [@huang2015predicting], our findings are backed by a large body of literature [??? Add REFs]. -->This suggested that the random forest model, when tuned and feature engineered correctly, is highly effective in predicting biologically relevant domain boundaries.

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values [??? So why didn't you use it?]. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling.

Guided by both low-resolution Hi-C data and high-resolution genome annotation data, *preciseTAD* predicts base-level resolution boundaries, alleviating resolution limitations of Hi-C data. We show that, unlike Arrowhead, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* TAD and loop boundaries were enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143, supporting recent observations that TADs and loops are dynamic structures generated by similar mechanisms [@hansen2018recent]. Likewise, *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, a known feature among the 3D architecture of the human genome, further highlighting their biological significance [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa].

*preciseTAD* offers flexibility in controlling both the number of TAD(loop) boundaries as well as the distance between them. The two primary parameters are the probability threshold (*t*) and $\epsilon$ (refered to as eps; parameter of DBSCAN). The combination of these two quantities change the resulting number of predicted boundaries from *preciseTAD*. A lower threshold and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 4 different thresholds ($t=(0.95,0.975,0.99,1.0)$) and 4 different $\epsilon$ values ($\epsilon=(0.2 r, 0.5r, r, 2r)$)), where $r=5000$, the optimal resolution. We found that the normalized overlaps--calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries --between top TFBS sites and flanked preciseTAD boundaries were maximized for combinations of t=1.0 and eps=r (Supplementary Figure 9). The same heuristic ($t=1.0$ and $\epsilon=r$) ensured that boundaries were at least 5kb bases away from each other, while also suitably enriched for architectural proteins (Supplementary Figures 10-11). These most optimal heuristics are set as default, but can be further tuned, if necessary.
